\section*{Acknowledgments}

I would like to express my sincere gratitude to all of the amazing people that have made my time in graduate school such a transformative and positive experience. I have had countless scientific, athletic, and personal adventures that wouldn't have been possible without all of you. There are so many people that have made a difference in my life during the last four and a half years, but in particular I would like to thank: 

\begin{itemize}

\item Yaniv Erlich, for taking me under his wing within weeks of starting his own lab, for being a terrific mentor, for sharing his brilliant creative energy, for giving me so many opportunities to grow as a scientist, and for making science so much fun.

\item My family: Alon Goren, for being an amazing husband and friend, as well as a great colleague. For
working together to build a dual-science career family, wholeheartedly supporting my own work and ambitions while pursuing his, for so many night and weekend science parties, and for putting up with my occasional whining. Eli Goren, for being the cutest most amazing child I could ever ask for, for putting up with marathon paper writing sessions in his first couple months of life, and for giving me renewed purpose and energy.

\item Mark Daly, for his great advising, first as a thesis committee member and then as a co-advisor during the last year of my PhD. His scientific and career input have been invaluable.

\item David Altshuler, for his valuable mentorship during several years of my PhD, and for his amazing ability to help me see the big picture of my work and career.

\item David Reich, for his scientific insight as a member of my thesis committee and a collaborator, and for all of his support during the last several years.

\item My labmates: Dina Zielinski, for putting up with me for four years while I listened to Michael Jackson, rap music, Disney songs, and Hanson in the lab, didn't listen to her through my headphones, talked to myself while programming, crossed the ``green line'' separating our desk spaces, and complained about how hungry I was. Thanks for being a great triathlon and science buddy, getting into trouble, and for being a great friend. Thomas Willems, for growing as scientists together and being such a great colleague. Assaf Gordon (Gordy) for responding to all of my programming questions with hours long monologues, for sharing his knowledge of just about any topic with us, and for making the lab a fun place to be. Mona Sheikh, for terming the term ``Broadents'' and helping Dina hack into my Amazon and Facebook to subscribe to rodent-lover groups. Joanna Kaplanis, for keeping the lab sane and for her amazing accent. Thank you also to Barak Markus, Mary Wahl, David Golan, and everyone else that has made the Erlich such a great home over the last several years.

\item Manolis Kellis and the students of 6.047 for helping me grow as a teacher.

\item Jason Wright, for his extreme generosity in introducing me to the world of experimental biology.

\item Leonid Mirny, for convincing me to come to HST and for his advice during the first year of my PhD.

\item All my fellow HST students that have shared in this experience with me. In particular Jenny Chen, Carolin Unglert, Laura Tannenbaum, Jesse Engreitz, Alal Eran, Eran Hodis.

\item All the caffeinated beverages that kept me going: ``Happy juice'' (caffeinated crystal lite), Area Four and The Study for their delicious coffee, and for the Whitehead and Broad vending machines. 

\item My running and triathlon friends: Zuzka Trnovcova, Kirstin Scott, Dina Zielinski, Isabel Brachmann, Justin Chen, and others who I have convinced to join me on ridiculously long running adventures, including 50K trail runs, and who have dragged me on their own cycling and swimming outings.

\item My parents, Dolly and Ken Gymrek, for giving me the best education possible and for supporting my goals. My sister, Megan Machucho, for giving me some space in the womb and sharing my genome, and of course for supporting me.





\end{itemize}